Adaptive evolution with aneuploidy and mutation

### 2022-03-23 ms/ms.tex
use macro for genotypes

%%% Table: fitness
\begin{table}[h]
\centering
\caption{\textbf{Fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^*$ & $2n^*$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:fitness}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Prior distributions.} We used informative prior distributions for $w_{2n+1}=1-c+b$, $w_{2n+1^*}=(1+s)(1-c)+b$ and $w_{2n^*}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.

We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$~(\Cref{fig:growth-curves}-D, assuming $w_{2n}=1$). In lieu of a more suitable prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
To increase computational efficiency, we also assumed $w(2n*)>w(2n+1*)>w(2n+1)>w(2n)$; running the inference without this assumption produced similar results. 
We also examined an extended informative prior that uses additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$~(\Cref{fig:growth-curves}L). The same prior was used for $w_{2n*}/w_{2n+1*}$.

As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$, or (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In these cases the inference algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
%% estimates generated by MAP-and-hdi.ipynb
The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower). YR: we need more than one ref; also what is the units of Zho 2014?
The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated fitness values are $w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.025\ [1.024-1.026]$,
$w_{2n^*}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$. 
Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:fitness}).

@@

Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). %% generated with MAP-and-hdi.ipynb
The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
The fitness of the mutant is also much lower at $w_{2n^*}=1.013\ [1.012-1.013]$.
This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations. Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).

@@

% generated with alt-prior.ipynb
Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with different priors for $w_{2n+1^*}$ and $w_{2n*}$. The estimated mutation rate was much lower compared to the basic prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.032\ [1.03-1.035]$,
$w_{2n^*}=1.033\ [1.031-1.036]$. 
The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}).

@@

% Fig Curveball 
\begin{figure}[h]
	\centering
	% generated with growth_curves.ipynb and deg = '39' 
	\includegraphics[width=.95\textwidth]{../figures/evo39_fitness_39deg.pdf} 
	% generated with growth_curves_refined.ipynb
	\includegraphics[width=0.95\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
\caption{
    \textbf{Fitness estimation from growth curves.}
    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{2n+1}/w_{2n}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $\hat w_{2n*}/w_{2n+1}$=1.033 (95\% CI: 1.027 - 1.041).
    Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
	Fitness estimated from growth curves using \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. See \emph{Models and Methods, Prior distributions} for more details.  \textbf{(A,B;E,F)} Mono-culture growth curve data (markers) and best-fit growth models (lines).
\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively, 6,375 generated curves. \textbf{(D,H)} The relative fitness distribution for \anwt\ relative to \euwt\ (panel D) and \eumt\ relative to \anwt\ (panel H). Figures generated by \texttt{Curveball}.
% TODO IK please fix the range of x axis in D; It definitely doesnt need to go beyond 1.25...
} 
\label{fig:growth-curves}
\end{figure}

@@

% Fig mu comparisons
%% generated with diff-mutation-rate.ipynb
\begin{figure}[h!]
  \centering
  \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/mu-A.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/mu-B.pdf}      
  \end{subfigure}
  \\
   \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/mu-C.pdf}      
  \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/mu-D.pdf}      
    \end{subfigure}
  \caption{
    \textbf{Model with fixed mutation rate.}  \textbf{(A-D)} The inferred posterior distributions for models with free and fixed mutation rate, $\mu$. The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are following: 
\textbf{free $\boldsymbol{\mu}$:}
$\delta=2.749\cdot10^{-3}\ [1.476\cdot10^{-3}-2.822\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.025\ [1.023-1.026]$,
$w_{2n^*}=1.027\ [1.026-1.029]$;
\textbf{free $\boldsymbol{\mu}$, run 2:}
$\delta=1.938\cdot10^{-3}\ [1.338\cdot10^{-3}-2.748\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.02-1.023]$,
$w_{2n+1^*}=1.025\ [1.023-1.026]$,
$w_{2n^*}=1.027\ [1.026-1.029]$;
\textbf{$\boldsymbol{\mu=10^{-5}}$:}
$\delta=3.089\cdot10^{-3}\ [2.412\cdot10^{-3}-3.169\cdot10^{-3}]$,
$w_{2n+1}=1.022\ [1.021-1.023]$,
$w_{2n+1^*}=1.024\ [1.023-1.026]$,
$w_{2n^*}=1.027\ [1.026-1.028]$;
\textbf{$\boldsymbol{\mu=10^{-6}}$:}
$\delta=1.413\cdot10^{-3}\ [1.04\cdot10^{-3}-2.529\cdot10^{-3}]$,
$w_{2n+1}=1.021\ [1.02-1.023]$,
$w_{2n+1^*}=1.024\ [1.023-1.026]$,
$w_{2n^*}=1.028\ [1.026-1.029]$;
\textbf{$\boldsymbol{\mu=10^{-7}}$:}
$\delta=3.4\cdot10^{-3}\ [2.043\cdot10^{-3}-3.578\cdot10^{-3}]$,
$w_{2n+1}=1.019\ [1.017-1.02]$,
$w_{2n+1^*}=1.026\ [1.024-1.027]$,
$w_{2n^*}=1.027\ [1.026-1.029]$.
\label{fig:mu}
  }
  \end{figure}

>>>>

%%% Table: fitness
\begin{table}[h]
\centering
\caption{\textbf{Fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & \euwt & \anwt & \anmt & \eumt \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:fitness}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Prior distributions.} We used informative prior distributions for $w_{\anwt}=1-c+b$, $w_{\anmt}=(1+s)(1-c)+b$ and $w_{\eumt}=1+s$, which we estimated from growth curves data from mono-culture growth experiments previously reported by \citet[Figs. 3C, 4A, and S2]{Yona2012}.
We used \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. Briefly, \texttt{Curveball} takes growth curves of two strains growing separately in mono-culture and predicts how they would grow in a mixed culture, that is, it predicts the results of a competition assay.
From these predictions, relative fitness values can be computed. Because \texttt{Curveball} uses a maximum-likelihood approach to estimate model parameters, we were able to estimate a distribution of relative fitness values by sampling from a truncated multivariate normal distribution defined by the maximum-likelihood covariance matrix.

We sampled 10,000 samples to use as a prior distribution~(\Cref{fig:growth-curves}).
We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{\anwt}$~(\Cref{fig:growth-curves}-D, assuming $w_{\euwt}=1$). In lieu of a more suitable prior, we used the same prior for $w_{\anmt}$ and $w_{\eumt}$. 
To increase computational efficiency, we also assumed $w(\eumt)>w(\anmt)>w(\anwt)>w(\euwt)$; running the inference without this assumption produced similar results. 
We also examined an extended informative prior that uses additional growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{\eumt}/w_{\anwt}$~(\Cref{fig:growth-curves}L). The same prior was used for $w_{\eumt}/w_{\anmt}$.

As a control, we tested an uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{\anwt}$, $w_{\anmt}$, $w_{\eumt}$, or (ii) only for $w_{\anmt}$, $w_{\eumt}$, using the above informative prior for $w_{\anwt}$. In these cases the inference algorithm failed to converge.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
%% estimates generated by MAP-and-hdi.ipynb
The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$, agrees with previous estimates. % TODO refs (9.7*10-5 in Zhu 2014, so it's lower). YR: we need more than one ref; also what is the units of Zho 2014?
The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:fitness}).

@@

Moreover, a model without aneuploidy (where the aneuploidy rate is fixed at zero, $\delta=0$), fails to explain the experimental observations~(Figure~\ref{fig:2n*-fixation}). %% generated with MAP-and-hdi.ipynb
The estimated mutation rate without aneuploidy is $\mu=7.979\cdot10^{-9}\ [7.903\cdot 10^{-9}-8.135\cdot10^{-9}]$, much lower compared to a model with aneuploidy and suggesting a target size of just 40. 
The fitness of the mutant is also much lower at $w_{\eumt}=1.013\ [1.012-1.013]$.
This is because, without aneuploidy, a high mutation rate or fitness effect will lead to faster appearance and fixation of \eumt\ than in the experimental observations. Even with these lower estimates, the model fit is worse than that of a model with aneuploidy~(\Cref{fig:2n*-fixation}).

@@

% generated with alt-prior.ipynb
Finally, we estimated the parameters under an extended informative prior (see \emph{Prior distributions}), with different priors for $w_{\anmt}$ and $w_{\eumt}$. The estimated mutation rate was much lower compared to the basic prior, with $\mu=2.379\cdot10^{-9}\ [2.267\cdot10^{-9}-2.458\cdot10^{-9}]$. Other parameter estimates are: $\delta=2.636\cdot10^{-3}\ [2.245\cdot10^{-3}-2.919\cdot10^{-3}]$,
$w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.032\ [1.03-1.035]$,
$w_{\eumt}=1.033\ [1.031-1.036]$. 
The WAIC of the model is much lower: -57 compared to 295~(\Cref{table:WAIC}).

@@

% Fig Curveball 
\begin{figure}[h]
	\centering
	% generated with growth_curves.ipynb and deg = '39' 
	\includegraphics[width=.95\textwidth]{../figures/evo39_fitness_39deg.pdf} 
	% generated with growth_curves_refined.ipynb
	\includegraphics[width=0.95\textwidth]{../figures/refined_vs_evo39_fitness_39deg.pdf}
\caption{
    \textbf{Fitness estimation from growth curves.}
    \textbf{(A-D)} Fitness estimation from growth curves of \euwt\ and \anwt\ at \SI{39}{\celsius}. $\hat w_{\anwt}/w_{\euwt}$=1.024 (95\% CI: 0.959 - 1.115). \texttt{Curveball} 
    \textbf{(E-H)} Fitness estimation from growth curves of \anwt\ and \eumt\ at \SI{39}{\celsius}. $\hat w_{\eumt}/w_{\anwt}$=1.033 (95\% CI: 1.027 - 1.041).
    Growth curves previously described in \citet[Figs. 3C, 4A, and S2]{Yona2012}.
	Fitness estimated from growth curves using \texttt{Curveball}, a method for predicting results of competition experiments from growth curve data~\citep[\href{https://curveball.yoavram.com}{curveball.yoavram.com}]{Ram2019}. See \emph{Models and Methods, Prior distributions} for more details.  \textbf{(A,B;E,F)} Mono-culture growth curve data (markers) and best-fit growth models (lines).
\textbf{(C,G)} The mixed-culture prediction for the strains from A,B and E,F respectively, 6,375 generated curves. \textbf{(D,H)} The relative fitness distribution for \anwt\ relative to \euwt\ (panel D) and \eumt\ relative to \anwt\ (panel H). Figures generated by \texttt{Curveball}.
% TODO IK please fix the range of x axis in D; It definitely doesnt need to go beyond 1.25...
} 
\label{fig:growth-curves}
\end{figure}

@@

% Fig mu comparisons
%% generated with diff-mutation-rate.ipynb
\begin{figure}[h!]
  \centering
  \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/mu-A.pdf}      
  \end{subfigure}
  \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/mu-B.pdf}      
  \end{subfigure}
  \\
   \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/mu-C.pdf}      
  \end{subfigure}
    \begin{subfigure}{0.45\textwidth}
      \includegraphics[width=\textwidth]{../figures/mu-D.pdf}      
    \end{subfigure}
  \caption{
    \textbf{Model with fixed mutation rate.}  \textbf{(A-D)} The inferred posterior distributions for models with free and fixed mutation rate, $\mu$. The MAP (maximum a posteriori) and 50\% HDI (highest density interval) for each model are following: 
\textbf{free $\boldsymbol{\mu}$:}
$\delta=2.749\cdot10^{-3}\ [1.476\cdot10^{-3}-2.822\cdot10^{-3}]$,
$w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.023-1.026]$,
$w_{\eumt}=1.027\ [1.026-1.029]$;
\textbf{free $\boldsymbol{\mu}$, run 2:}
$\delta=1.938\cdot10^{-3}\ [1.338\cdot10^{-3}-2.748\cdot10^{-3}]$,
$w_{\anwt}=1.022\ [1.02-1.023]$,
$w_{\anmt}=1.025\ [1.023-1.026]$,
$w_{\eumt}=1.027\ [1.026-1.029]$;
\textbf{$\boldsymbol{\mu=10^{-5}}$:}
$\delta=3.089\cdot10^{-3}\ [2.412\cdot10^{-3}-3.169\cdot10^{-3}]$,
$w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.024\ [1.023-1.026]$,
$w_{\eumt}=1.027\ [1.026-1.028]$;
\textbf{$\boldsymbol{\mu=10^{-6}}$:}
$\delta=1.413\cdot10^{-3}\ [1.04\cdot10^{-3}-2.529\cdot10^{-3}]$,
$w_{\anwt}=1.021\ [1.02-1.023]$,
$w_{\anmt}=1.024\ [1.023-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$;
\textbf{$\boldsymbol{\mu=10^{-7}}$:}
$\delta=3.4\cdot10^{-3}\ [2.043\cdot10^{-3}-3.578\cdot10^{-3}]$,
$w_{\anwt}=1.019\ [1.017-1.02]$,
$w_{\anmt}=1.026\ [1.024-1.027]$,
$w_{\eumt}=1.027\ [1.026-1.029]$.
\label{fig:mu}
  }
  \end{figure}
